---
title: "Conditional, Loops, and Functions"
output: html_notebook
---


## This lecture is about

- conditionals (if/else)
- loops (for, while)
- how to define your own functions
- basic programming flow



# Conditionals

## if statements


```{r}


```


---  

```{r}

```

## if/else statements

```{r}
# Flip a coin 

```


## else-if statements

```{r}

```


## Equality

```{r}

```

## Beware of finite precision arithmetic

```{r}
# oops
sqrt(2)^ 2 ==2
```

_safety first_

```{r}
dplyr::near(sqrt(2)^2, 2)
```

## Vectorized operations

```{r}

```

## Use the double symbol in for loops

default to `&&` or `||` in a for loop

```{r}

```

Otherwise you get a vector

```{r}

```

# Loops

## for loops

```{r}
for (i in 1:10) {
    print(i)
}


```

## Pre-allocate memory 

```{r}
nums <- vector("double", 10)
for (i in 1:10) {
  nums[i] <- runif(1)
}

nums


```

## Dynamic allocation bad

```{r}
# nums <- c()
# for (i in 1:10) {
#   nums <- c(nums, runif(1))
# }
```


## while loops


```{r}
current_position <- 10
n_iter <- 0
while (current_position >0) {
    current_position <- current_position + rnorm(1)
    n_iter <- n_iter +1
}
print(paste0('You lost all your money after ', n_iter, ' Trips to the casino'))
```


## Infinite loops

```{r}
# while (TRUE) {
#     print('Math rocks')
# }


```

## Vectorization

Try to vectorize anything you can 

```{r}

sapply(1:10, function(x)x * 2)

```

```{r}
n <-10
sum((1:n)^2)
test <- (1:10)^2
sum(test)
```

```{r}
sum(c(NA, 1,2,-3, pi), na.rm = T)
```

# Functions

## Functions are for humans and computers

- break long program up into small chunks
- more readable
- code reuse
    - catch errors
    - don't have to copy/paste
- only need to fix code in one place

## When to write a function?


> You should consider writing a function whenever you've copied and pasted a block of code more than twice

## Define a function

```{r}
power <- function(num, exponent) {
    num ^ exponent
}
power(2,4)
```

## Default arguments

```{r}

power <- function(num, exponent=3) {
    num ^ exponent
}
power(3)

```

## Return values

```{r}
# random_rps <- function() {
#     r <- runif(1) 
#         return('rock')
#     }else if (1/3 < r && r< 2/3){
#         return('paper')
#     }else{
#         return('Scissors')
#     }
```


## Modularity and abstraction

- _modularity_
    - break a complicated task into many smaller sub-tasks
    
- _abstraction_
    - you don't need to  know the inner workings of every part of the system, just how things interact
    


# A note on vectors


## Vector vs List

- vectors are homogeneous and sequential (1 dimensional)

- lists are heterogeneous and hierarchical 

## Vectors have a type

- boolean

- character

- complex

- raw

- integer 

- double


## Integer vector

```{r}
int <- 1:10
int
```

## Boolean vector

```{r}
# boolean
bool <- c(T,F,T)

```

## Character vector

```{r}
# string
caract <- c('a','b','c')

```


## `typeof()`

```{r}
typeof(caract)

```



## Question 1

What are the types of the following vectors?

```{r eval=F}
# a
c(1, 2, 'three')

# b
c(TRUE, TRUE, "FALSE")

# c
c(1, 2, 3.1)
```

## Explicit coercion


```{r}

as.integer(c('1','2','3'))

```

## Implicit coercion

```{r}



```

then 

```{r}

```


## `rbind()` and `cbind()`


```{r}
u <- c(20,3,pi)
v <- c(-1,0, exp(1))
v

```


```{r}
rbind(u,v)

```

```{r}

```



## Coercion

- Logical values are converted to numbers: TRUE is converted to 1 and FALSE to 0.

- Values are converted to the simplest type required to represent all information.

- The ordering is roughly:

logical < integer < numeric < complex < character < list.

- Object attributes are dropped when an object is coerced from one type to another.



## Vectorized operations and implicit coercion

```{r}
c(-2,-1,1,2,3) >0

sum(c(-2,-1,1,2,3)>0)

```

## Subsetting a vector

```{r}
v <- 10:20
v [c(1,10)]

v[v %% 2 == 0]


```

```{r}
a <- c(2,3,4)
b <- c(1,2)

```

# A note on lists

## Lists

Lists can contain objects of multiple types and are indexed by names (as opposed to index sequentially)

Like a vector, a list is an indexed set of objects (and so has a length), but unlike a vector the elements of a list can be of different types, including other lists! The _mode_ of a list is `list`.

## Lists

- A list is just a generic container for other objects and the power and utility of lists comes from this generality. 

- A list might contain an individual measurement, a vector of observations on a single response variable, a dataframe, or even a list of dataframes containing the results of several experiments. 

- 

## Make a list

```{r}

```

## Access elements of a list with `[[]]`

Single square brackets are used to select a sublist; double square brackets are used to extract a single element.


To access elements of a list use `[[]]` 

```{r}

```

## A single bracket returns a list


you can use a single `[]` and this will return a list 

```{r}


```

see [R4DS](http://r4ds.had.co.nz/vectors.html#lists) for the difference between `[]` and `[[]]`


## Lists are hierarchical

```{r}

```


```{r}

```

## Another example

