---
title: "Subsetting"
date: "Adapted from \"Advanced R\" by Hadley Wickham"
output: html_notebook
---




## Before we jump into it...

Make sure that you have R and Rstudio installed in your computer

_Alternatively:_ feel free to use [rstudio.cloud](https://rstudio.cloud) for a web-based version of RStudio

**Good references**

- _"R for Data Science"_ (R4DS) by Hadley Wickham and Garret Grolemund. The official website of the textbook is <http://r4ds.had.co.nz/>

- "Advanced-R" <http://adv-r.had.co.nz/>







# Atomic vectors

---

```{r}


```

There are five things that you can use to subset a vector:

- Positive integers return elements at the specified positions:

```{r}


```

```{r}


```

```{r}
# Duplicated indices yield duplicated values


```


---

- Negative integers omit elements at the specified positions:

```{r}


```

- Logical vectors select elements where the corresponding logical value is `TRUE`. 

```{r}


```


```{r}


```

---

- Nothing returns the original vector. This is not useful for vectors but is very useful for matrices, data frames, and arrays. 

```{r}


```


- Zero returns a zero-length vector. This is not something you usually do on purpose, but it can be helpful for generating test data.

```{r}


```

---

If the vector is named, you can also use:

```{r}


```

```{r}


```


# Lists

---

Subsetting a list works in the same way as subsetting an atomic vector. 

Using `[` will always return a list; `[[` and `$` let you pull out the components of the list.

## Matrices and arrays

Supply a 1d index for each dimension, separated by a comma. Blank subsetting is now useful because it lets you _keep all rows or all columns_.

```{r}



```


```{r}


```

## Data frames

Data frames possess the characteristics of both lists and matrices: if you subset with a single vector, they behave like lists; if you subset with two vectors, they behave like matrices.

```{r}


```

```{r}


```

---

```{r}
# Like a list:


```

```{r}
# Like a matrix

```

# Subsetting operators

---


`[[` is similar to `[`, except it can only return a single value and it allows you to pull pieces out of a list. 

You need `[[` when working with lists. This is because when `[` is applied to a list it always returns a list: it never gives you the contents of the list. To get the contents, you need `[[`:


```{r}


```


```{r}


```


Because data frames are lists of columns, you can use `[[` to extract a column from data frames: `mtcars[[1]]`, `mtcars[["cyl"]]`.


## `$`

`$` is a shorthand operator, where `x$y` is equivalent to `x[["y", exact = FALSE]]`. It is often used to access variables in a data frame, as in `mtcars$cyl` or `diamonds$carat`.

- One important difference between `$` and `[[`: `$` does partial matching

```{r}


```


```{r}


```

---


```{r}


```


## Missing/out of bounds indices

`[` and `[[` differ slightly in their behaviour when the index is out of bounds (OOB), for example, when you try to extract the fifth element of a length four vector, or subset a vector with `NA` or `NULL`:

```{r}



```



```{r}


```

```{r}


```


# Some Applications of Subsetting

---

## Lookup tables

Character matching provides a powerful way to make lookup tables. Say you want to convert abbreviations:

```{r}
x <- c("m", "f", "u", "f", "f", "m", "m")
lookup <- c(m = "Male", f = "Female", u = NA)
lookup[x]
```

```{r}
#don't want names in the result: use unname() to remove 
unname(lookup[x])
```

```{r}
# Or with fewer output values
c(m = "Known", f = "Known", u = "Unknown")[x]
```


## Matching and merging by hand 

Suppose we have a vector of integer grades, and a table that describes their properties:

```{r}
grades <- c(1, 2, 2, 3, 1, 3, 3)

info <- data.frame(
  grade = 3:1,
  desc = c("Excellent", "Good", "Poor"),
  fail = c(F, F, T)
)
```

<small>
We want to duplicate the info table so that we have a row for each value in `grades`. We can do this in two ways, either using `match()` and integer subsetting, or `rownames()` and character subsetting:
</small>

```{r}
grades
```

---

```{r}
# Using match
id <- match(grades, info$grade)
info[id, ]
```



```{r}
# Using rownames
rownames(info) <- info$grade
info[as.character(grades), ]
```

## Expanding aggregated counts 

Sometimes you get a data frame where identical rows have been collapsed into one and a count column has been added. `rep()` and integer subsetting make it easy to uncollapse the data by subsetting with a repeated row index:

```{r}
df <- data.frame(x = c(2, 4, 1), y = c(9, 11, 6), n = c(3, 5, 1))
df
```


```{r}
rep(1:nrow(df), df$n)
```

---

```{r}
df[rep(1:nrow(df), df$n), ]
```

## Removing columns from data frames 

```{r}
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])
df$z <- NULL
```

```{r}
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])
df[c("x", "y")]
```

## Selecting rows based on a condition

```{r}


```

```{r}

```


Use the vector boolean operators `&` and `|`, not the _short-circuiting_ scalar operators `&&` and `||` which are more useful inside if statements. 


## `subset()`

```{r}


```

