---
title: "Data Import"
output: html_notebook
---

We will use tools from the `readr` package that is part of the `tidyverse`

```{r, message=FALSE}
library(tidyverse)
```



Refer to the "Data Import" chapter from [R4DS](https://r4ds.had.co.nz/data-import.html) for additional information.

- `read_csv()` reads comma delimited files

- `read_csv2()` reads semicolon separated files (common in countries where , is used as the decimal place)

- `read_tsv()` reads tab delimited files

- `read_delim()` reads in files with any delimiter (specified with the option `delim`)

We practice by reading a `.csv` file directly from the web:

```{r}
# create a string variable with the location of the file
nba_link <- "https://raw.githubusercontent.com/reisanar/datasets/master/NBAchampionsdata.csv"

# read the file using the read_csv() function
nba_champions <- read_csv(nba_link)
```

The file contains information regarding the final games for the NBA champions of every year.


The first argument to `read_csv()` is the most important: the path to the file to read. When you run `read_csv()` it prints out a column specification that gives the name and type of each column.



Suppose we want to check the distribution of points per game (variable `PTS`). We can use a histogram to explore that variable

```{r}
nba_champions %>%
  ggplot(aes (x = PTS)) +
  geom_histogram()

```

It seems there are some NBA champions that scored more than 130 points in a finals game:

```{r}

filter(nba_champions, PTS > 130)

```

```{r}
filter(nba_champions, PTS > 100, Year == 2001)
```

Some datasets come with missing values, denoted as `NA` in R.

```{r}
# define location of file
rating_link <- "https://raw.githubusercontent.com/reisanar/datasets/master/courserating.csv"
# read courses ratings data with read_csv()
course_ratings <- read_csv(rating_link)
```

If we print the data frame we can observe the missing values

```{r}
head(course_ratings)

```

```{r}
course_ratings %>%
  select(everything()) %>%
  summarise_all(funs(sum(is.na(.))))
```


Some other items to keep in mind:

- Sometimes there are a few lines of metadata at the top of the file. You can use `skip = n` to skip the first `n` lines; or use `comment = "#"` to drop all lines that start with (e.g.) `#`.

- The data might not have column names. You can use `col_names = FALSE` to tell `read_csv()` not to treat the first row as headings, and instead label them sequentially from `X1` to `Xn`

