---
title: "Homework Assignment 1"
subtitle: "Data Exploration and Preprocessing"
author: "Danilo Cardoso dos Santos - dcardoso8563@floridapoly.edu"
output: html_notebook
---

## Problem 1

Load the `tidyverse` and the `nycflights13` packages.

```{r}
library(tidyverse)
library(nycflights13)
```

Using tools from `dplyr` (part of the `tidyverse`), answer the following questions:


(a) From the list of flights that left NYC in 2013 (using the `flights` data frame), find those with a departure delay larger than 60 minutes

```{r}
flights %>%
  filter(year == 2013 & dep_delay > 60)

```


(b) How would you find the flights with a scheduled departure time later than 10:30 PM ? 

```{r}
# 2130 == 10:30 PM

flights %>%
  filter(sched_dep_time > 2130)
```



(c) Find all flights that were operated by United, American, or Delta.

```{r}
# United   == UA
# American == AA
# Delta    == DL

flights %>%
  filter(carrier %in% c('UA','AA','DL'))

```


## Problem 2

In this problem you will be using tools from the `caret` package (short for **C**lassification **A**nd **RE**gression **T**raining)

```{r}
# load the caret package
library(caret)
```

(a) Consider the data from the `mtcars` data frame (a built-in dataset in R): 

```{r}
# print the last 6 entries in the mtcars data frame
tail(mtcars)

```


(b) Using tools from the `caret` package (discussed in class), perform the following transformations to the `mtcars` data frame: **normalization**, **standardization**, and **centering**.

Comments on your results. Verify that in the standardization process, the transformed variables have _unit standard deviation_ (Hint: the `sd()` function in R computes the standard deviation). Choose 3 variables to confirm this. 


**Normalizing**

```{r}
preProc_norm <- preProcess(mtcars, method = c("range"))

new_norm <- predict(preProc_norm, mtcars)

summary(new_norm)

```


## Standardizing

```{r}
preProc_std <- preProcess(mtcars, method = c("center", "scale"))

new_std <- predict(preProc_std, mtcars)

summary(new_std)

# Verifying the Standard Deviation of the new Processed DS

# All three tested variables resulted in 1 for the SD

sd(new_std$mpg)

sd(new_std$cyl)

sd(new_std$disp)

```


## Centering

```{r}

preProc_center <- preProcess(mtcars, method = "center")

new_center <- predict(preProc_center, mtcars)

## Checking the new Summarized values
summary(new_center)

```


## Problem 3

The chunk of code below, defines a function in R to create a Hilbert matrix. 

```{r}
# create Hilbert matrix of size n 
hilbert <- function(n) { 
  i <- 1:n
  1 / outer(i - 1, i, "+") 
  }
```


(a) Build a Hilbert matrix of size 7, and call it `hil_seven`. 

```{r}

hil_seven <- hilbert(7)

```


(b) Select columns 1 through 4, and call it `X`

```{r}
X <- hil_seven[,1:4] 

```


(c) Compute the **singular value decomposition** of `X` using the `svd()` function.

```{r}

s <-svd(X)
```


(d) Print the singular values of `X`

```{r}
s$d
```


(e) Verify that the product $U^T U$ where $U$ is the matrix of left singular values, returns the _identity matrix_ (a square matrix with ones in the main diagonal, and zeros everywhere else). Recall that the `t()` function in R, creates the _transpose_ of a matrix.

```{r}
myU <- s$u
```


```{r}
t(myU) %*% myU
```

