---
title: "Homework Assignment 2"
author: "Danilo Cardoso dos Santos `dcardoso@floridapoly.edu`"
subtitle: Introduction to R programming
output: html_notebook
---


# Problem 1
(adapted from spuRs, chapter 3, exercise 6)

Write a program in R in which given a vector $x$, its geometric mean is calculated. Recall that the _geometric mean_ of $x_1, . . . , x_n$ is 

$$
\left(\prod_{i=1}^{n}x_{i}\right)^{\frac{1}{n}} 
$$

```{r}
x <- c(3, 4)

gm <- 1

for (xi in x)gm <-gm*xi
  (gm <- gm^(1/length(x)))

```



Create a second function that calculates the _harmonic mean_

$$
\frac{n} {\left(\sum_{i=1}^{n}\frac{1}{x_{i}}\right)} = \left(\frac{\sum_{i=1}^n x_i^{-1}}{n} \right)^{-1}
$$

(notice, from the second form, that the harmonic mean is simply the reciprocal of the _arithmetic mean_ of reciprocals)
```{r}
hm <- 1/sum(1/x)

hm
```


- Checks that if the $x_i$ are all positive, the harmonic mean is always less than or equal to the geometric mean (which is always less than or equal to the arithmetic mean).

```{r}


for (i in length(x)) {
  if(xi>0){
    print("All Positives")
  }else{
    print("Not all positives")
  }
}


```

# Problem 2 
(adapted from spuRs, chapter 3, exercise 7)

Write a program in R that finds the sum of every third element of a vector `x`.

```{r}
x <- c(1,2,100,7,5,99,88)

third <- seq(3,length(x),by=3)

sum(x[third])

```

# Problem 3 

Is `x[-which(x > 0)]` the same as `x[x <= 0]`? Show some examples.

(see the documentation for `which()` as needed)

```{r}
x <- c(1,2,0,-1,-3)

x[-which(x > 0)]

x[x <= 0]

```


# Problem 4

Read the documentation for the function `upper.tri()`. What does `upper.tri()` return? How does subsetting a matrix with it work? Provide an example

```{r}
#help(upper.tri)

M <- matrix(1:20,4, 5)

UT<-upper.tri(M)

M[UT]

```

_It returns an upper triangular of a matrix_

_It can return the TRUE values of an upper_


# Problem 5

Write a function that calculates the mean of a numeric vector `x`, ignoring the `s` smallest and `l` largest values (this is a **_trimmed mean_**).

For example, if `x = c(1, 7, 3, 2, 5, 0.5, 9, 10)`, `s = 1`, and `l = 2`, your function would return the mean of `c(1, 7, 3, 2, 5)` (this is `x` with the 1 smallest value (0.5) and the 2 largest values (9, 10) removed).

Your function should use the `length()` function to check if `x` has at least `s + l + 1` values. If `x` is shorter than `s + l + 1`, your function should use the `message()` function to tell the user that the vector cannot be trimmed as requested. If `x` is at least length `s + l + 1`, your function should return the trimmed mean.

_Hint_: For this exercise you might use the `sort()` function.

Define your function with 3 possible inputs:

```{r}
# Informative comments describing what the function does

x = c(1, 7, 3, 2, 5, 0.5, 9, 10)

trimmedMean <- function(x, s=0, l=0) {
  if(length(x)< s+l+1){
    message("The vector cannot be trimmed")
  }else{
  sorted_a <- sort(x)
  sorted_d <- sort(x, decreasing=T)
  ss <- sorted[(1:s)]
  ll <- sorted_d[(1:l)]  
  sl <-c(ss,ll)
  mean(x[!x%in%sl])
    
  }
}
trimmedMean(x,1,2)

```

Note that the `s = 0` and `l = 0` specified in the function definition are the _default settings_. i.e., this syntax ensures that if `s` and `l` are not provided by the user, they are both set to 0. Thus the default behavior is that the `trimmedMean` function does not trim anything, and hence is the same as the `mean` function.
``


# Problem 6

Consider the following function

```{r}
to.string <- function(x) {
  string <- x[1]
  for(i in 2:length(x)) {
    string <- paste(string, x[i], sep="")
  }
  return(string)
}
```

- What does the above function `to.string` do to a vector `u`?

```{r}
u <- c(1,2,3,4)

to.string(u)

```

_It returns a vector with multiple elements to just one element converted to string_

- What is the output of `to.string(2:17)` ? 

```{r}

to.string(2:17)

```


- Can you think of a more elegant or simpler implementation of that function? _Hint_ : look at the help page for the `R` function `paste`

```{r}
# simpler way
paste(c(u),collapse='')

```


# Problem 7

Create a 6 × 10 matrix of random integers chosen from 1, 2,. . . , 10 by executing the following two lines of code:

```{r}
set.seed(217)
aMat <- matrix( sample(10, size=60, replace=T), nr=6)
```


(a) Use R to find the number of entries in each row which are greater than 4.
```{r}
  
coun <- ifelse(aMat > 4,1,0)

rowSums(coun)

```

(b) Which rows contain _exactly two_ occurrences of the number seven?
```{r}

coun <- ifelse(aMat == 7,1,0)

rowSums(coun)

```


